---
title: "OGM Feature Model Analysis"
format:
  html:
    toc: true
    embed-resources: true
execute:
  warning: false
  message: false
---

## 1. Setup

This document only visualizes and statistically tests data - it does not run any simulations itself. Simulation logic and data generation live in `analysis.py`; run it first (`python analysis.py`) to populate `data/*.csv`, then re-render this document whenever the data changes.

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

AGENT_COLORS <- c(FeatureAgent = "#1f77b4", Agent = "#ff7f0e")
AGENT_LABELS <- c(FeatureAgent = "SF", Agent = "SV")

params_df <- read_csv("data/params.csv", show_col_types = FALSE)
params <- setNames(params_df$value, params_df$name)

ALPHA <- as.numeric(params["ALPHA"])
GAMMA <- as.numeric(params["GAMMA"])
TEMP <- as.numeric(params["TEMP"])
VALUE_DECAY <- as.numeric(params["VALUE_DECAY"])
TRAUMA_VALENCE <- as.numeric(params["TRAUMA_VALENCE"])
N_RETRIEVALS <- as.integer(params["N_RETRIEVALS"])
MAX_STEPS <- as.integer(params["MAX_STEPS"])
N_REPS <- as.integer(params["N_REPS"])
SWEEP_REPS <- as.integer(params["SWEEP_REPS"])
decay_rate_values <- as.numeric(strsplit(gsub("\\[|\\]", "", params["decay_rate_values"]), ",\\s*")[[1]])
```

## 2. Agent vs FeatureAgent Comparison

Direct comparison of the single-value `Agent` against the per-feature-value `FeatureAgent`: a single-run illustration with trauma-encounter markers (`single_run.png`), and the original network with one fixed trauma node from initialization (`comparison_1.png`). Data comes from `data/single_run.csv` and `data/comparison_1.csv` (generated by `analysis.py`). The statistical-testing helpers defined here are reused in section 3.

```{r}
sig_stars <- function(p) {
  case_when(
    p < 0.001 ~ "***",
    p < 0.01  ~ "**",
    p < 0.05  ~ "*",
    TRUE      ~ "ns"
  )
}

# Welch's t-test (unequal variance) between Agent and FeatureAgent at each
# param value, for each dv_spec (list with dv_name, agent_raw, feature_raw;
# *_raw is a list of per-rep numeric vectors, one per param value).
# Holm-Bonferroni correction (base R's p.adjust) is applied across every row
# of the returned table.
agent_vs_feature_table <- function(param_values, param_name, dv_specs) {
  rows <- list()
  for (spec in dv_specs) {
    for (i in seq_along(param_values)) {
      test <- t.test(spec$agent_raw[[i]], spec$feature_raw[[i]])
      rows[[length(rows) + 1]] <- data.frame(
        param_value = param_values[i],
        dependent_variable = spec$dv_name,
        agent_mean = mean(spec$agent_raw[[i]]),
        featureagent_mean = mean(spec$feature_raw[[i]]),
        t = unname(test$statistic),
        p = test$p.value
      )
    }
  }
  df <- do.call(rbind, rows)
  df$p_holm_adj <- p.adjust(df$p, method = "holm")
  df$sig <- sig_stars(df$p_holm_adj)
  names(df)[names(df) == "param_value"] <- param_name
  df
}

# One-way ANOVA (equal-variance, matching classic F-test) across all param
# values, for each agent_spec (list with agent_name, dv_name, raw_by_value;
# raw_by_value is a list of per-rep numeric vectors, one per param value).
# Holm-Bonferroni correction is applied across every row of the returned table.
across_values_table <- function(agent_specs) {
  rows <- list()
  for (spec in agent_specs) {
    long <- data.frame(
      value = unlist(spec$raw_by_value),
      group = factor(rep(seq_along(spec$raw_by_value), lengths(spec$raw_by_value)))
    )
    test <- oneway.test(value ~ group, data = long, var.equal = TRUE)
    rows[[length(rows) + 1]] <- data.frame(
      agent = spec$agent_name,
      dependent_variable = spec$dv_name,
      F = unname(test$statistic),
      p = test$p.value
    )
  }
  df <- do.call(rbind, rows)
  df$p_holm_adj <- p.adjust(df$p, method = "holm")
  df$sig <- sig_stars(df$p_holm_adj)
  df
}

# Pairwise post-hoc Welch's t-tests between adjacent param_values (i.e.
# consecutive levels only, not every combination), for each agent_spec (as in
# across_values_table). These follow up the omnibus ANOVA by localizing which
# specific adjacent pair(s) of param values differ.
# Holm-Bonferroni correction is applied separately within each agent x
# dependent-variable group of pairwise comparisons, since each group answers
# a distinct question and shouldn't dilute another group's error budget.
pairwise_across_values_table <- function(param_values, param_name, agent_specs) {
  groups <- list()
  for (spec in agent_specs) {
    idx <- seq_along(param_values)
    pairs <- lapply(head(idx, -1), function(i) c(i, i + 1))
    group_rows <- list()
    for (pr in pairs) {
      i <- pr[1]; j <- pr[2]
      test <- t.test(spec$raw_by_value[[i]], spec$raw_by_value[[j]])
      row <- data.frame(
        agent = spec$agent_name,
        dependent_variable = spec$dv_name,
        a = param_values[i],
        b = param_values[j],
        t = unname(test$statistic),
        p = test$p.value
      )
      names(row)[names(row) == "a"] <- paste0(param_name, "_A")
      names(row)[names(row) == "b"] <- paste0(param_name, "_B")
      group_rows[[length(group_rows) + 1]] <- row
    }
    group_df <- do.call(rbind, group_rows)
    group_df$p_holm_adj <- p.adjust(group_df$p, method = "holm")
    group_df$sig <- sig_stars(group_df$p_holm_adj)
    groups[[length(groups) + 1]] <- group_df
  }
  do.call(rbind, groups)
}
```

```{r}
single_run_df <- read_csv("data/single_run.csv", show_col_types = FALSE)

feature_1 <- single_run_df %>% filter(agent_type == "FeatureAgent", retrieval_num <= 250) %>% arrange(retrieval_num)
agent_1 <- single_run_df %>% filter(agent_type == "Agent", retrieval_num <= 250) %>% arrange(retrieval_num)
feature_1$cumulative_reward <- cumsum(feature_1$reward)
agent_1$cumulative_reward <- cumsum(agent_1$reward)

single_run_long <- bind_rows(feature_1, agent_1)

plot_df <- single_run_long %>%
  pivot_longer(cols = c(length, cumulative_reward), names_to = "metric", values_to = "value") %>%
  mutate(
    metric = case_when(
      metric == "length" ~ "Retrieval Length",
      metric == "cumulative_reward" ~ "Cumulative Reward"
    ),
    agent_type = factor(agent_type, levels = c("FeatureAgent", "Agent"))
  )

ggplot(plot_df, aes(x = retrieval_num, y = value)) +
  geom_line(aes(color = agent_type), show.legend = FALSE) +
  facet_grid(metric ~ agent_type, scales = "free_y", labeller = labeller(agent_type = AGENT_LABELS)) +
  scale_color_manual(values = AGENT_COLORS) +
  labs(x = "Retrieval number", y = NULL) +
  theme_minimal()

ggsave("plots/single_run.png", width = 12, height = 8, dpi = 150)
```

```{r}
comparison_df <- read_csv("data/comparison_1.csv", show_col_types = FALSE)

comparison_df <- comparison_df %>%
  arrange(agent_type, rep, retrieval_num) %>%
  group_by(agent_type, rep) %>%
  mutate(cumulative_reward = cumsum(reward)) %>%
  ungroup()

comparison_summary <- comparison_df %>%
  group_by(agent_type, retrieval_num) %>%
  summarise(
    length_mean = mean(length),
    length_sem = if (n() > 1) sd(length) / sqrt(n()) else 0,
    .groups = "drop"
  ) %>%
  mutate(agent_type = factor(agent_type, levels = c("FeatureAgent", "Agent")))

ggplot(comparison_summary, aes(x = retrieval_num, y = length_mean, color = agent_type, fill = agent_type)) +
  geom_ribbon(aes(ymin = length_mean - length_sem, ymax = length_mean + length_sem), alpha = 0.2, color = NA) +
  geom_line() +
  scale_color_manual(values = AGENT_COLORS, labels = AGENT_LABELS) +
  scale_fill_manual(values = AGENT_COLORS, labels = AGENT_LABELS) +
  labs(x = "Retrieval number", y = "Average Retrieval Length over Time", color = NULL, fill = NULL) +
  theme_minimal()

ggsave("plots/comparison.png", width = 12, height = 5, dpi = 150)
```

## 3. Decay Rate Sweep (Agent vs FeatureAgent)

Compares average retrieval length across `decay_rate` - 0 (no decay), 0.01, 0.02 - for both `Agent` and `FeatureAgent` on the original network (`decay_rate_1.png`). `decay_rate_values` matches Kato & Morita's (2016, *PLOS Comp Biol* e1005145) explored phi/kappa no-decay-to-decay range; `ALPHA`/`GAMMA`/`TEMP` stay at their global defaults (loaded from `data/params.csv`) throughout. Data comes from `data/decay_rate_sweep.csv`. Cumulative reward is intentionally excluded here - it tracks the same underlying effect as retrieval length and is redundant with it.

```{r}
decay_df <- read_csv("data/decay_rate_sweep.csv", show_col_types = FALSE)

raw_by_decay_rate <- function(df, agent) {
  df %>%
    filter(agent_type == agent) %>%
    group_by(decay_rate) %>%
    summarise(values = list(avg_retrieval_length), .groups = "drop") %>%
    arrange(match(decay_rate, decay_rate_values)) %>%
    pull(values)
}

agent_raw_avg_len <- raw_by_decay_rate(decay_df, "Agent")
feature_raw_avg_len <- raw_by_decay_rate(decay_df, "FeatureAgent")

decay_summary <- decay_df %>%
  group_by(agent_type, decay_rate) %>%
  summarise(
    mean_length = mean(avg_retrieval_length),
    sem_length = if (n() > 1) sd(avg_retrieval_length) / sqrt(n()) else 0,
    .groups = "drop"
  ) %>%
  mutate(agent_type = factor(agent_type, levels = c("FeatureAgent", "Agent")))

ggplot(decay_summary, aes(x = decay_rate, y = mean_length, color = agent_type)) +
  geom_errorbar(aes(ymin = mean_length - sem_length, ymax = mean_length + sem_length), width = 0.001) +
  geom_line() +
  geom_point() +
  scale_color_manual(values = AGENT_COLORS, labels = AGENT_LABELS) +
  labs(x = "decay_rate", y = "Retrieval length", color = NULL) +
  theme_minimal()

ggsave("plots/decay_rate.png", width = 6, height = 5, dpi = 150)
```

Statistical tests accompanying the plot above:

1. Agent vs FeatureAgent at each decay_rate (Welch's t-test, unequal variance).
2. Across decay_rate values, within each agent type (one-way ANOVA, 3 levels).
3. Pairwise post-hoc comparisons between adjacent decay_rate values (0 vs 0.01, 0.01 vs 0.02), within each agent, to localize which specific pair(s) the ANOVA picked up on.

Only average retrieval length is tested (cumulative reward dropped - see the sweep cell above). Every table's p-values are Holm-Bonferroni corrected (`p.adjust(method = "holm")`) within that table - see each heading for exactly which set of tests share a correction (the family-wise error budget).

### Agent vs FeatureAgent, at each decay_rate

Welch's t-test, Holm-adjusted across all 3 tests:

```{r}
agent_vs_feature <- agent_vs_feature_table(
  decay_rate_values, "decay_rate",
  list(list(dv_name = "avg retrieval length", agent_raw = agent_raw_avg_len, feature_raw = feature_raw_avg_len))
)
kable(agent_vs_feature, digits = 4)
```

### Pairwise decay_rate comparisons, within each agent

Welch's t-test, Holm-adjusted within each agent's group of 2 pairwise tests:

```{r}
pairwise_decay <- pairwise_across_values_table(
  decay_rate_values, "decay_rate",
  list(
    list(agent_name = "Agent", dv_name = "avg retrieval length", raw_by_value = agent_raw_avg_len),
    list(agent_name = "FeatureAgent", dv_name = "avg retrieval length", raw_by_value = feature_raw_avg_len)
  )
)
kable(pairwise_decay, digits = 4)
```
